---
title: "Ridge Implementation"
author: "Karan Gill"
date: "2023-08-08"
output: html_document
---
Experimental Design: 

In the experiment below the dataset being considered is a Player Valuation dataset. The dataset contains 50 features about 519 players from the top 5 European Leagues Premier League (England), La Liga (Spain), Serie A (Italy), Bundesliga (Germany) and Ligue 1(France). The experiment aims to calculate the beta coefficient for the Ridge Regression Model using the method of the formula. We have also calculated the optimal lambda which minimizes the mean squared error in the Ridge Regression Model in this scenario and used that value of lambda to calculate the coefficients of the model.

The dataset has certain players which have missing values for their columns. These players have been omitted from the dataset on which the regression is being performed for the sake of simplicity.

```{r}
# Load necessary libraries
library(readxl)
library(dplyr)
library(glmnet)

# Load the dataset
setwd("C:/Users/Karan/Desktop/Dissertation")
data <- read_excel("Player_Valuation.xlsx")

# Remove rows with missing values
data_clean <- na.omit(data)

# Separate predictors and response
X <- data_clean %>% select(-Ranking, -Player, -Market_Value)
y <- data_clean$Market_Value

# Encode categorical variables
X_encoded <- model.matrix(~., data = X)

# Perform cross-validation for Ridge Regression
set.seed(123) # For reproducibility
cv_fit <- cv.glmnet(X_encoded, y, alpha=0)  # alpha=0 for Ridge Regression

# Extract the optimal lambda value
optimal_lambda <- cv_fit$lambda.min

# Print the optimal lambda
optimal_lambda
```

The optimal value of Lambda is 3.47551. For this value of lambda the mean squared error is minimized. This value of lambda is used to calculate the coefficients of the model using the formula in R.

```{r}
# Compute Ridge Regression coefficients using the formula for the optimal lambda
beta_ridge <- solve(t(X_encoded) %*% X_encoded + optimal_lambda * diag(ncol(X_encoded))) %*% t(X_encoded) %*% y

# Print the coefficients for reference
beta_ridge
```

The Intercept represents the response variable Market_Value. When the other variables have no value (0) it is the estimated value of the Market_Value which is 2.16329628 (in millions). 

The other coefficients represent the change in the value of Market_Value variable for a one-unit change in the predictor, assuming all other predictors are held constant. For binary predictors (such as one-hot encoded categorical variables like Nation, Pos, etc), the coefficient represents the difference in Market_Value between the various categories.